Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-07-24 pituba/article/article_spe.tex
more and more

\begin{abstract}
This work examines the underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations. The three-dimensional model is integrated in reservoir volume and quantities such as fluid properties are spatially averaged. The material balance equation is obtained for the averaged quantities and it is applied to calculate reservoir pressure. It is shown that the calculated pressure corresponds to the local pressure at a defined depth and not to the average reservoir pressure in general. The definition of a convenient reference depth is discussed. Sufficient conditions for solution uniqueness of the material balance equation are provided for the case that pressure is the unknown quantity and one of such conditions is shown to check for physical correctness of PVT properties.
Finally, a scheme for solving the material balance equation numerically is proposed.
\end{abstract}

@@

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can thus be defined. Similarly, a pseudo oil component can also be defined by subtracting all the pseudo gas component from oil phase composition. After defining these two pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.

@@

We assume that the reservoir is a connected volume where hydrocarbon saturation is nonzero. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases. Gas phase composition is always that of a dry gas.

@@

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume, i.e.,
\begin{equation}
T_w = T_o = T_g = T_{rock} = T = \text{constant} \, .
\end{equation}

Temperature is a constant prescribed value and thus no equation for energy balance is needed. It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if the component number is restricted to the two pseudo-components. As temperature is assumed constant, oil phase density is a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:

@@

To close the system of equations, a connection among different phase pressures is provided by capillary pressure relations:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\vec{\vec{k}},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\vec{\vec{k}},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

@@

All quantities are initially functions of space and time, but they become functions of time only after being averaged in space:

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at surface condition, i.e.:

@@

\subsubsection{Relationship between the material balance equation and the reservoir pressure}

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by the following expression:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
\end{equation}
where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.
It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.

It can be shown that neglecting the variation of fluid properties with depth does not introduce any important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.

@@

An open source implementation of secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.

@@

Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e., holding all produced or injected volumes constant and allowing only pressure to change. Each different solid line corresponds to a different volume of fluid production and injection. The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure. Also, it is seen that function $\epsilon$ monotonically decreases with pressure, as theoretically shown before.

Figure~\ref{fig: iter} shows the count of newtonian iterations required to compute pressure with an accuracy within $0.001$ of initial bubble-point pressure. Typically, $12$ iterations are required.

>>>>

\begin{abstract}
This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations. The three-dimensional model is integrated in reservoir volume and quantities such as fluid properties are spatially averaged. The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure. It is shown that the calculated pressure corresponds to the local pressure at an arbitrary reference depth and not to the average reservoir pressure in general. The definition of a convenient reference depth is discussed. Sufficient conditions for solution uniqueness of the material balance equation are provided for the case that pressure is the unknown quantity. One of such conditions provides a check for physical correctness of PVT properties obtained from laboratory experiments or empirical correlations.
Finally, a numerical scheme for solving the material balance equation is proposed.
\end{abstract}

@@

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo-gas component can be defined. Similarly, a pseudo-oil component can be defined by subtracting all the pseudo-gas component from oil phase composition. After performing this procedure, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.

@@

We assume that the reservoir is a connected volume where hydrocarbon saturation is nonzero. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases, i.e., gas phase composition is always that of a dry gas.

@@

Another simplification is now made assuming thermal equilibrium in the reservoir volume, i.e.,
\begin{equation}
T_w = T_o = T_g = T_{rock} = T = \text{constant} \, .
\end{equation}

Temperature is a constant prescribed value and no equation for energy balance is needed. It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if the component number is restricted to the two pseudo-components. As temperature is assumed constant, oil phase density is a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:

@@

To close the system of equations, a connection among different phase pressures is provided by capillary pressure relations:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}=p_g-p_o \, .
\end{equation}

@@

All quantities are initially functions of space and time, but they become functions of time alone after being averaged in space:

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at surface condition, i.e.:

@@

\subsubsection{Relationship between the material balance equation and reservoir pressure}

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth given by the following expression:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
\end{equation}
where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.
It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.

It can be shown that neglecting the variation of density with depth does not introduce any important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.

@@

An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.

@@

Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) when pressure is varying and all remaining terms are held constant. Each different solid line corresponds to a different instant of time and therefore different volumes of cumulative production and injection. The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure. Also, it is seen that function $\epsilon$ monotonically decreases with pressure, as theoretically shown before.

Figure~\ref{fig: iter} shows the count of newtonian iterations required to compute pressure with an accuracy within $0.1\%$ of the initial bubble-point pressure. Typically, $12$ iterations are required.
